---
title: "p8105_hw3_brm2150"
author: "Brooklynn McNeil"
date: "2024-10-07"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(p8105.datasets)
library(ggridges)
library(patchwork)

knitr::opts_chunk$set(
  comment = '', fig.width = 8, fig.height = 6, out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## Problem 1

Clean the `ny_noaa` data by separating the variables for year, month, and day. Ensure units for temp, precipitation, and snowfall. What is the most commonly observed values for snowfall and why?
```{r}
data("ny_noaa")

ny_noaa_clean = 
  ny_noaa |>
  filter(!if_any(prcp:tmin, is.na)) |>
  separate(date, into = c("year", "month", "day"), sep = "-") |>
  mutate(
    prcp = prcp/254, 
    snow = snow/25.4, 
    snwd = snwd/25.4,
    tmax = as.numeric(tmax),
    tmax = ((tmax*0.1*9/5) + 32),
    tmin = as.numeric(tmin),
    tmin = ((tmin*0.1*9/5) + 32)
    )
```

Below is a two-panel plot showing the average max temperature in January and in July in each station across years. 
```{r}
hex = 
  ny_noaa_clean |> 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_hex()

ridge = 
  ny_noaa_clean |> 
  filter(snow < 100, snow > 0) |>
  ggplot(aes(x = snow, y = as.factor(year))) + 
  geom_density_ridges()

hex + ridge
```


## Problem 2

Import and tidy accelerometer data.
```{r}
nhanes_accel = 
  read_csv("data/nhanes_accel.csv") |>
  janitor::clean_names()|>
  pivot_longer(
    cols = starts_with("min"),
    names_to = "minute",
    names_prefix = "min",
    values_to = "MIMS")

nhanes_covar = 
  read_csv("data/nhanes_covar.csv", skip = 4) |>
  janitor::clean_names()

nhanes_tidy = 
  left_join(nhanes_accel, nhanes_covar, by = "seqn")|>
  filter(age>=21)|>
  drop_na()|>
  mutate(
    sex = case_when(
     sex == 1 ~ "male",
      sex == 2 ~ "female"),
    sex = factor(sex),
    education = case_when(
      education == 1 ~ "Below HS Equivalent",
      education == 2 ~ "HS Equivalent",
      education == 3 ~ "Above HS"),
    education = factor(education),
    age = factor(age),
    seqn = factor(seqn)
     )
```
Reader friendly table for observing the number of men and women in each education category. Individuals with an above high school education for both sexes are more active throughout the day. What is odd about the data is that the group with the least amount of activity is females with a high school education. 

```{r}
nhanes_tidy|>
  group_by(sex, education) |>
  summarize(
    count = n()
  ) |>
  arrange(group = desc(count)) |>
  knitr::kable()
```
```{r echo=TRUE, message=FALSE, warning=FALSE}
nhanes_tidy |>
  distinct(seqn, .keep_all = TRUE) |>
  ggplot(aes(x = age, fill = sex)) +
  geom_density(alpha = 0.5) +
  scale_x_discrete(breaks = c(20,30,40,50,60,70,80,90,100)) +
    facet_grid(~factor(education, levels = c("Below HS Equivalent","HS Equivalent", "Above HS")))
```

Here we analyze the activity data from the accelerometers. First, I pivoted wider so that I could create a total activity variable. Below is a plot comparing the total activity among the variables `age`, `sex`, and `education`. From the plots below, we can see that women with at least a high school education are more active than their male counter parts at almost all ages. For people with high school equivalent education, the activity peaks at age 40 for women and about 35 for men, and then continues to decline as age increases. 

```{r}
nhanes_tidy |>
  pivot_wider(
    names_from = minute,
    values_from = MIMS,
    names_prefix = "min") |>
  mutate(total_activity = rowSums(across(.cols = c(min1:min1440))))|>
  relocate(seqn, sex,age,bmi,education,total_activity) |>
  ggplot(aes(x = age, y = total_activity, color = sex)) +
  geom_point(alpha = 0.5) + 
  geom_smooth(aes(group = sex), se = FALSE) +
  scale_x_discrete(breaks = c(20,30,40,50,60,70,80,90,100)) +
  facet_grid(~factor(education, levels = c("Below HS Equivalent","HS Equivalent", "Above HS"))) +
  labs(
    title = "Total Activity from Accelerometers"
  )
```
Make a graph with 3 panels to display 24-hour activity for each educational level and color to indicate sex. Add smoothing trends
I made a three-panel plot to display the 24-hour data that was collected by the accelerometers. From the plots below, we can tell that most people start to get up and become active around and become less active around 7 or 8 o'clock. Again, the women with at least a high school degree show more activity throughout the entire day than the men. Both sexs with less than a High School education seem to have a peak activity around 10 in the morning and then slowly decline activiity throughout the day. 
```{r}
nhanes_tidy |>
  mutate(time_of_day = as.numeric(minute)/60)|>
  ggplot(aes(x = time_of_day, y = MIMS, color = sex)) +
  geom_point(alpha = 0.01) +
  geom_smooth(aes(group = sex))+
  scale_y_continuous(limits = c(1,60))+
  facet_grid(~factor(education, levels = c("Below HS Equivalent","HS Equivalent", "Above HS"))) +
  labs(
    title = "24-Hour Activity from Accelerometers"
  )
```


## Problem 3

Import and tidy citi bike data.
```{r}
jan_2020 = 
  read_csv("data/citibike/Jan 2020 Citi.csv") |>
  mutate(month = "january",
         year = "2020")

jan_2024 = 
  read_csv("data/citibike/Jan 2024 Citi.csv")|>
  mutate(month = "january",
         year = "2024")

jul_2020 = 
  read_csv("data/citibike/July 2020 Citi.csv")|>
  mutate(month = "july",
         year = "2020")

jul_2024 = 
  read_csv("data/citibike/July 2024 Citi.csv")|>
  mutate(month = "july",
         year = "2024")

citi_bike_df = 
  bind_rows(jan_2020, jan_2024, jul_2020, jul_2024)
```

Below is a table counting the number of rides that casual riders and members take in the months of January and July and the years 2020 and 2024.The table reflects that citi bike members clearly ride more often than casual riders. It also shows that Citi bike has been getting more popular overtime as well as being more popular in July than in Janurary.  

```{r}
citi_bike_df |>
  group_by(month, year, member_casual) |>
  summarize(count = n()) |>
  arrange(count)|>
  knitr::kable()
```

Below is a table of the most popular stations to start a ride in July 2024.

```{r}
citi_bike_df |>
  filter(month == "july", year == 2024) |>
  group_by(month, year, start_station_name) |>
  summarize(count = n()) |>
  arrange(desc(count)) |>
  slice(1:5) |>
  knitr::kable(label = "Top 5 Start Station in July 2024")
```

Day of the week, month and year are being investigated for their effect on median ride duration. The data is very skewed, so I removed outliers from the boxplot. Rides that occur in July tend to be longer than those in January. Interestingly, ride duration has decreased from 2020 to 2024. Not so suprisngly, the ride duration is longer on Saturdays and Sunday. Looks like people are enjoying their weekend rides!

```{r}
weekday = 
  citi_bike_df |>
  ggplot(aes(x = duration, y = weekdays, fill = weekdays)) +
  geom_boxplot(outliers = FALSE)

month = 
  citi_bike_df |>
  ggplot(aes(x = duration, y = month, fill = month)) +
  geom_boxplot(outliers = FALSE)

year= 
  citi_bike_df |>
  ggplot(aes(x = duration, y = year, fill = year)) +
  geom_boxplot(outliers = FALSE)

(month + year)/weekday
```

For data in 2024, make a figure that shows the impact of month, membership status, and bike type on the distribution of ride duration. Comment on your results.

The data from 2024 includes more information from the electric bikes. The boxplots below show that the median ride duration is much more affected by the bike type when it is a casual ride and not one from a member. 

```{r}
citi_bike_df |>
  filter(year == 2024)|>
  ggplot(aes(x = rideable_type, y = duration, fill = rideable_type)) +
  geom_boxplot(outliers = FALSE)+
  facet_grid(~month ~member_casual)
```

