---
title: "p8105_hw3_brm2150"
author: "Brooklynn McNeil"
date: "2024-10-07"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(p8105.datasets)

knitr::opts_chunk$set(
  comment = '', fig.width = 5, fig.height = 5
)
```

## Problem 1

Clean the `ny_noaa` data by separating the variables for year, month, and day. Ensure units for temp, precipitation, and snowfall. What is the most commonly observed values for snowfall and why?
```{r}
data("ny_noaa")

ny_noaa_clean = 
  ny_noaa |>
  filter(!if_any(prcp:tmin, is.na)) |>
  separate(date, into = c("year", "month", "day"), sep = "-") |>
  mutate(
    prcp = prcp/254, 
    snow = snow/25.4, 
    snwd = snwd/25.4,
    tmax = as.numeric(tmax),
    tmax = ((tmax*0.1*9/5) + 32),
    tmin = as.numeric(tmin),
    tmin = ((tmin*0.1*9/5) + 32)
    )
```

Make a two-panel plot showing the average max temperature in January and in July in each station across years. Is there any observable / interpretable structure? Any outliers?
```{r}
#ny_noaa_clean |>
 # filter(month == c("01", "07")) |>
 #   mutate(
 #   month = case_match(month, "01" ~ "January", "07" ~ "July")
 # ) |>
  #ggplot(aes(x = day, y = tmin, color = year)) +
  #geom_point(alpha = 0.2) +
  #geom_smooth(se = FALSE, na.rm = TRUE) +
  #facet_grid(~month)
```


## Problem 2

Import and tidy accelerometer data.
```{r}
nhanes_accel = 
  read_csv("data/nhanes_accel.csv") |>
  janitor::clean_names()|>
  pivot_longer(
    cols = starts_with("min"),
    names_to = "minute",
    names_prefix = "min",
    values_to = "MIMS")

nhanes_covar = 
  read_csv("data/nhanes_covar.csv", skip = 4) |>
  janitor::clean_names()

nhanes_tidy = 
  left_join(nhanes_accel, nhanes_covar, by = "seqn")|>
  filter(age>=21)|>
  drop_na()|>
  mutate(
    sex = case_when(
     sex == 1 ~ "male",
      sex == 2 ~ "female"),
    sex = factor(sex),
    education = case_when(
      education == 1 ~ "Below equivalent",
      education == 2 ~ "HS equivalent",
      education == 3 ~ "Above HS"),
    education = factor(education),
    minute = factor(minute),
    age = factor(age),
    seqn = factor(seqn)
     )
```
Reader friendly table for observing the number of men and women in each education category. 

```{r}
nhanes_tidy|>
  group_by(sex, education) |>
  summarize(
    count = n()
  ) |>
  knitr::kable()
```


Create a visualization of the age distributions for men and women in each educational category. Comment on these
```{r echo=TRUE, message=FALSE, warning=FALSE}
nhanes_tidy |>
  distinct(seqn, .keep_all = TRUE) |>
  select(age, sex, education)|>
  ggplot(aes(x = age, fill = sex)) +
  geom_density(alpha = 0.5) +
  scale_x_discrete(breaks = c(20,30,40,50,60,70,80,90,100)) +
  facet_grid(~education)
```



## Problem 3

Import and tidy citi bike data.
```{r}
jan_2020 = 
  read_csv("data/citibike/Jan 2020 Citi.csv") |>
  mutate(month = "january",
         year = "2020")

jan_2024 = 
  read_csv("data/citibike/Jan 2024 Citi.csv")|>
  mutate(month = "january",
         year = "2024")

jul_2020 = 
  read_csv("data/citibike/July 2020 Citi.csv")|>
  mutate(month = "july",
         year = "2020")

jul_2024 = 
  read_csv("data/citibike/July 2024 Citi.csv")|>
  mutate(month = "july",
         year = "2024")

citi_bike_df = 
  bind_rows(jan_2020, jan_2024, jul_2020, jul_2024)
```

Describe the data

```{r}
citi_bike_df |>
  group_by(duration, weekdays) |>
  summarize()

citi_bike_df |>
  ggplot(aes(x = weekdays, duration, fill = member_casual)) +
  geom_boxplot()
```


